---
title: "Home Credit Default Risk — Modeling"
author: "Rob Sheehy"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    df_print: paged
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This notebook builds predictive models for the Home Credit Default Risk project. Based on the EDA findings, we focus on logistic regression models and explore handling the class imbalance issue.

# Load Data
```{r}
library(dplyr)
library(readr)

# Load the original application_train.csv instead
train_data <- read_csv("application_train.csv", show_col_types = FALSE)

cat("Data dimensions:", nrow(train_data), "rows,", ncol(train_data), "columns\n\n")

# Check TARGET exists
cat("TARGET column check:\n")
print(table(train_data$TARGET))
cat("Default rate:", round(100 * mean(train_data$TARGET), 2), "%\n\n")

# Select key features for modeling
train_clean <- train_data %>%
  select(TARGET, 
         EXT_SOURCE_1, EXT_SOURCE_2, EXT_SOURCE_3,
         AMT_CREDIT, AMT_INCOME_TOTAL, AMT_ANNUITY,
         DAYS_BIRTH, DAYS_EMPLOYED) %>%
  mutate(
    AGE_YEARS = abs(DAYS_BIRTH) / 365,
    CREDIT_INCOME_RATIO = AMT_CREDIT / AMT_INCOME_TOTAL,
    ANNUITY_INCOME_RATIO = AMT_ANNUITY / AMT_INCOME_TOTAL
  ) %>%
  select(-DAYS_BIRTH, -DAYS_EMPLOYED, -AMT_CREDIT, -AMT_INCOME_TOTAL, -AMT_ANNUITY) %>%
  na.omit()

cat("After cleaning and feature engineering:", nrow(train_clean), "observations\n")
cat("Features:", ncol(train_clean) - 1, "\n")
```

# Train/Validation Split
```{r}
library(caret)

set.seed(42)

# Create 80/20 split
train_index <- createDataPartition(train_clean$TARGET, p = 0.8, list = FALSE)

train_set <- train_clean[train_index, ]
valid_set <- train_clean[-train_index, ]

cat("Training set:", nrow(train_set), "observations\n")
cat("Training default rate:", round(100 * mean(train_set$TARGET), 2), "%\n\n")

cat("Validation set:", nrow(valid_set), "observations\n")
cat("Validation default rate:", round(100 * mean(valid_set$TARGET), 2), "%\n")
```

# Baseline Logistic Regression
```{r}
library(yardstick)
library(pROC)

# Fit baseline model
baseline_model <- glm(TARGET ~ ., 
                     data = train_set, 
                     family = binomial())

# Get predictions
train_pred <- predict(baseline_model, type = "response")
valid_pred <- predict(baseline_model, newdata = valid_set, type = "response")

# Calculate AUC
train_auc <- roc_auc_vec(truth = factor(train_set$TARGET), 
                         estimate = train_pred, 
                         event_level = "second")

valid_auc <- roc_auc_vec(truth = factor(valid_set$TARGET), 
                         estimate = valid_pred, 
                         event_level = "second")

cat("Baseline Logistic Regression:\n")
cat("Training AUC:", round(train_auc, 4), "\n")
cat("Validation AUC:", round(valid_auc, 4), "\n\n")

# Feature importance
coef_df <- data.frame(
  feature = names(coef(baseline_model))[-1],
  coefficient = coef(baseline_model)[-1]
) %>%
  arrange(desc(abs(coefficient)))

knitr::kable(head(coef_df, 10), digits = 4, 
             caption = "Top 10 Feature Coefficients")

# ROC Curve
roc_obj <- roc(valid_set$TARGET, valid_pred, quiet = TRUE)
plot(roc_obj, main = "ROC Curve - Baseline Model",
     col = "#4682B4", lwd = 2)
abline(a = 0, b = 1, lty = 2, col = "gray")
text(0.6, 0.2, paste("AUC =", round(valid_auc, 3)), cex = 1.2)
```

# Weighted Logistic Regression
```{r}
# Calculate class weights to handle imbalance
class_weights <- 1 / table(train_set$TARGET)
weights <- ifelse(train_set$TARGET == 1, 
                  class_weights["1"], 
                  class_weights["0"])
weights <- weights * nrow(train_set) / sum(weights)

# Fit weighted model
weighted_model <- glm(TARGET ~ .,
                     data = train_set,
                     family = binomial(),
                     weights = weights)

# Get predictions
train_pred_w <- predict(weighted_model, type = "response")
valid_pred_w <- predict(weighted_model, newdata = valid_set, type = "response")

# Calculate AUC
train_auc_w <- roc_auc_vec(truth = factor(train_set$TARGET),
                           estimate = train_pred_w,
                           event_level = "second")

valid_auc_w <- roc_auc_vec(truth = factor(valid_set$TARGET),
                           estimate = valid_pred_w,
                           event_level = "second")

cat("Weighted Logistic Regression:\n")
cat("Training AUC:", round(train_auc_w, 4), "\n")
cat("Validation AUC:", round(valid_auc_w, 4), "\n\n")

cat("Improvement over baseline:", 
    round((valid_auc_w - valid_auc) * 100, 2), "percentage points\n")
```

# Model Comparison
```{r}
# Compare ROC curves
roc_baseline <- roc(valid_set$TARGET, valid_pred, quiet = TRUE)
roc_weighted <- roc(valid_set$TARGET, valid_pred_w, quiet = TRUE)

plot(roc_baseline, col = "#4682B4", lwd = 2, main = "Model Comparison")
plot(roc_weighted, col = "#CD5C5C", lwd = 2, add = TRUE)
abline(a = 0, b = 1, lty = 2, col = "gray")
legend("bottomright", 
       legend = c(paste("Baseline AUC =", round(valid_auc, 3)),
                  paste("Weighted AUC =", round(valid_auc_w, 3))),
       col = c("#4682B4", "#CD5C5C"), 
       lwd = 2)

# Summary table
results <- data.frame(
  Model = c("Baseline Logistic", "Weighted Logistic"),
  Train_AUC = c(train_auc, train_auc_w),
  Valid_AUC = c(valid_auc, valid_auc_w)
)

knitr::kable(results, digits = 4, caption = "Model Performance Summary")
```

# Summary

THe Logistic regression models demonstrate solid predictive performance on this imbalanced dataset. The baseline model achieves reasonable AUC scores using key features identified in exploratory analysis. The weighted approach attempts to address class imbalance by adjusting for the 8% default rate. EXT_SOURCE variables emerge as the most important predictors, consistent with domain knowledge about credit risk. Future work could explore additional feature engineering, threshold optimization for business objectives, and comparison with tree-based ensemble methods.